---
title: "Tidy Eval in R<br/>for Munging UDS Data"
author: "Nicolas May"
date: "July 31, 2019"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(dplyr)
library(knitr)
library(kableExtra)

pretty_print <- function(df) {
  df %>% 
    kable(align = "c") %>% 
    kable_styling(
      font_size = 16,
      bootstrap_options = c("striped", "bordered", "hover", "condensed", 
                            "responsive"))
}

pretty_print_scroll <- function(df) {
  df %>% 
    kable(align = "c") %>% 
    kable_styling(
      font_size = 16,
      bootstrap_options = c("striped", "bordered", "hover", "condensed", 
                            "responsive")) %>% 
    scroll_box(width = "100%")
}
```

# Tidy Eval

## Tidy Eval in R

_**Tidy Eval**_ is a conceptual framework for doing metaprogramming in R. 

> - **Metaprogramming** is the idea that we can treat code as data -- data that can be acted on by other code. 
> - When we treat **code as data**, it can be read, written, analyzed, and edited by other code.
> - Writing the "other code" that handles code-as-data is **metaprogramming**.

<br />
<p style="font-size:14px;">For a deep dive into this topic, check out the Metaprogramming section in [_**Advanced R**_](https://adv-r.hadley.nz/) (2nd edition) by Hadley Wickham.</p>

# Setup

<!-- ## EPA Fuel Economy Data -->

```{r epa_mpg, echo=FALSE, eval=TRUE}
# library(readxl)
# epa_mpg <- read_excel("./19tstcar-2019-03-14.xlsx") %>% 
#   select(make  = `Represented Test Veh Make`,
#          model = `Represented Test Veh Model`,
#          trim  = `Test Veh Configuration #`,
#          displ = `Test Veh Displacement (L)`,
#          type  = `Vehicle Type`,
#          hp    = `Rated Horsepower`,
#          cyl   = `# of Cylinders and Rotors`,
#          trans = `Tested Transmission Type`,
#          gears = `# of Gears`,
#          whls  = `Drive System Description`,
#          wght  = `Equivalent Test Weight (lbs.)`,
#          mpg   = `RND_ADJ_FE`,
#          categ = `Test Category`) %>% 
#   filter(trim == 0, categ == "FTP") %>% 
#   group_by()
```

## `mtcars` as a Tibble

```{r mtcars_1, echo=TRUE, eval=TRUE}
library(dplyr); library(stringr)
mtcars_tbl <- mtcars %>% 
  as_tibble(rownames = "model") %>% 
  mutate(model = case_when(
    str_detect(model, "Hornet") ~ paste("AMC", model),
    str_detect(model, "Camaro") ~ paste("Chevrolet", model),
    str_detect(model, "Duster") ~ paste("Plymouth", model),
    str_detect(model, "Valiant") ~ paste("Plymouth", model),
    str_detect(model, "Merc") ~ 
      str_replace(model, "Merc", "Mercedes"),
    TRUE ~ model
  )) %>% 
  mutate(make = str_replace(model, "(\\w+)\\W.+", "\\1")) %>% 
  mutate(model = str_replace(model, paste0(make, " "), "")) %>% 
  mutate_at(c("cyl", "hp", "vs", "am", "gear", "carb"), as.integer) %>% 
  select(make, model, everything())
```

## `mtcars` as a Tibble

```{r mtcars_2, echo=TRUE, eval=TRUE}
mtcars_tbl %>% head()
```

## `mtcars` as a Tibble

```{r mtcars_3, echo=TRUE, eval=TRUE}
mtcars_tbl %>% glimpse()
```

# Expressions

## Expressions as Arguments

> - To treat code as data, we need some way to capture the code expressions before they're evaluated.

> - Those familiar with the R [tidyverse](https://www.tidyverse.org/) have seen this in action with `dplyr`.

```{r expressions_1, echo=TRUE, eval=TRUE}
library(dplyr)
filter(mtcars_tbl, hp > 300) # hp = horsepower
```

> - Notice the filter condition `hp > 300` isn't a string. It's an expression.

## Expressions vs. Strings

> - `hp > 300` is an expression.

> - `"hp > 300"` is a string.

```{r expressions_2, echo=TRUE, eval=TRUE}
library(dplyr)
filter(mtcars_tbl, hp > 300) # hp = horsepower
```

> - How does `dplyr` do this? How does it capture the expression?

## Expression Capture

Suppose we'd like to build our own function that helps us summarize a data frame (like `mtcars_tbl`) with some descriptive statistics.

Here's a simple example with `dplyr::summarize` to build on:

```{r how_dplyr_1, echo=TRUE, eval=TRUE}
mtcars_tbl %>% 
  summarize(mean_hp = mean(hp))
```

## Expression Capture

Instead of finding the mean of all cars, what if we want to group the cars by number of cylinders, `cyl`?

```{r how_dplyr_2, echo=TRUE, eval=TRUE}
mtcars_tbl %>% 
  ### <b>
  group_by(cyl) %>% 
  ### </b>
  summarize(mean_hp = mean(hp))
```

## Expression Capture

Of course, we could group by other fields like `gear`.

```{r how_dplyr_3, echo=TRUE, eval=TRUE}
mtcars_tbl %>% 
  ### <b>
  group_by(gear) %>% 
  ### </b>
  summarize(mean_hp = mean(hp))
```

## Expression Capture

Or `make`.

```{r how_dplyr_4, echo=TRUE, eval=TRUE}
mtcars_tbl %>% 
  ### <b>
  group_by(make) %>% 
  ### </b>
  summarize(mean_hp = mean(hp)) %>% head()
```

## Expression Capture

Can we create a custom function that allows us to pass whatever grouping variable we want (`cyl`, `gear`, `make`)?

```{r how_dplyr_5, echo=TRUE, eval=TRUE}
group_by_mean_hp <- function(df, group_var) {
  df %>% 
    ### <b>
    group_by(group_var) %>% 
    ### </b>
    summarize(mean_hp = mean(hp))
}
```

Call the function using `cyl` as a grouping variable.

```{r how_dplyr_6, echo=TRUE, eval=TRUE, error=TRUE}
mtcars_tbl %>% 
  group_by_mean_hp(cyl)
```

## Expression Capture

Within the function, `group_var` is unknown. We'd have to add it explicitly _**within**_ the function for it to work. But that defeats the purpose.

```{r how_dplyr_7, echo=TRUE, eval=TRUE}
group_by_mean_hp <- function(df, group_var) {
  df %>% 
    ### <b>
    group_by(cyl) %>% 
    ### </b>
    summarize(mean_hp = mean(hp))
}
mtcars_tbl %>% 
  group_by_mean_hp(cyl)
```

## Expression Capture

How do we capture the `cyl` expression passed to `group_by_mean_hp` by the user?

We need to use tidy eval, specifically the `enexpr` function from the `rlang` package. `enexpr` captures the expression passed by the user.

```{r how_dplyr_8, echo=TRUE, eval=TRUE}
library(rlang)
group_by_mean_hp <- function(df, group_var) {
  ### <b>
  group_var_expr <- enexpr(group_var)
  ### </b>
  df %>% 
    ### <b>
    group_by(group_var_expr) %>%
    ### </b>
    summarize(mean_hp = mean(hp))
}
```

## Expression Capture

Let's give the updated `group_by_mean_hp` function a whirl.

```{r how_dplyr_10, echo=TRUE, eval=TRUE, error=TRUE}
mtcars_tbl %>% 
  group_by_mean_hp(cyl)
```

There's still an error. Why?

## Expression Capture

We "quoted" (or captured) the expression `cyl`. 

```{r how_dplyr_11a, echo=TRUE, eval=FALSE}
group_by_mean_hp <- function(df, group_var) {
  group_var_expr <- enexpr(group_var)
  df %>% 
    ### <b>
    group_by(group_var_expr) %>% 
    ### </b>
    summarize(mean_hp = mean(hp))
}
```

For `group_by` to evaluate it, "unquote" it with `!!` operator.

```{r how_dplyr_11b, echo=TRUE, eval=TRUE}
group_by_mean_hp <- function(df, group_var) {
  group_var_expr <- enexpr(group_var)
  df %>% 
    ### <b>
    group_by(!!group_var_expr) %>% 
    ### </b>
    summarize(mean_hp = mean(hp))
}
```

## Expression Capture

Let's try again with `cyl`.

```{r how_dplyr_12, echo=TRUE, eval=TRUE, error=TRUE}
mtcars_tbl %>% 
  group_by_mean_hp(cyl)
```

## Expression Capture

And again with `gear`.

```{r how_dplyr_13, echo=TRUE, eval=TRUE, error=TRUE}
mtcars_tbl %>% 
  group_by_mean_hp(gear)
```

## Expanding on Expression Capture

Can we generalize `group_by_mean_hp` a bit? 

What if we want the mean of something other than `hp`?

Well, we can use the same principles we just applied to the `group_var` argument.

```{r expand_expr_1, echo=TRUE, eval=TRUE}
group_by_mean <- function(df, group_var, mean_var) {
  group_var_expr <- enexpr(group_var)
  mean_var_expr <- enexpr(mean_var)
  
  df %>% 
    group_by(!!group_var_expr) %>% 
    summarize(mean = mean(!!mean_var_expr))
}
```

## Expanding on Expression Capture

Let's try it out.

```{r expand_expr_2, echo=TRUE, eval=TRUE}
mtcars_tbl %>% 
  group_by_mean(group_var = cyl, mean_var = hp)
```

But `mean` as a summary table label isn't very meaningful. 

Can we improve on this?

## Expanding on Expression Capture

```{r expand_expr_3, echo=TRUE, eval=FALSE, error=FALSE}
group_by_mean <- function(df, group_var, mean_var) {
  group_var_expr  <- enexpr(group_var)
  mean_var_expr   <- enexpr(mean_var)
  
  ### <b>
  mean_label_str  <- paste0("mean_", as_string(mean_var_expr))
  mean_label_expr <- enexpr(mean_label_str)
  ### </b>
  
  df %>% 
    group_by(!!group_var_expr) %>% 
    ### <b>
    summarize(!!mean_label_expr = mean(!!mean_var_expr))
  ### </b>
}
```

## Expanding on Expression Capture

```{r expand_expr_4, echo=TRUE, eval=FALSE, error=FALSE}
group_by_mean <- function(df, group_var, mean_var) {
  group_var_expr  <- enexpr(group_var)
  mean_var_expr   <- enexpr(mean_var)
  
  ### <b>
  mean_label_str  <- paste0("mean_", as_string(mean_var_expr))
  mean_label_expr <- enexpr(mean_label_str)
  ### </b>
  
  df %>% 
    group_by(!!group_var_expr) %>% 
    ### <b>
    summarize(!!mean_label_expr = mean(!!mean_var_expr))
  ### </b>
}
```

```{r expand_expr_5, echo=FALSE, eval=TRUE, error=TRUE}
group_by_mean <- function(df, group_var, mean_var) {
  group_var_expr  <- enexpr(group_var)
  mean_var_expr   <- enexpr(mean_var)
  
  mean_label_str  <- paste0("mean_", as_string(mean_var_expr))
  mean_label_expr <- enexpr(mean_label_str)
  
  df %>% 
    group_by(!!group_var_expr) %>% 
    summarize(!!mean_label_expr = mean(!!mean_var_expr))
}
```

## Expanding on Expression Capture

```{r expand_expr_6, echo=TRUE, eval=TRUE}
group_by_mean <- function(df, group_var, mean_var) {
  group_var_expr  <- enexpr(group_var)
  mean_var_expr   <- enexpr(mean_var)
  
  ### <b>
  mean_label_str  <- paste0("mean_", as_string(mean_var_expr))
  mean_label_expr <- enexpr(mean_label_str)
  ### </b>
  
  df %>% 
    group_by(!!group_var_expr) %>% 
    ### <b>
    summarize(!!mean_label_expr := mean(!!mean_var_expr))
  ### </b>
}
```

When the LHS is an unquoted expression, we need `:=`, a special assignment operator.

## Expanding on Expression Capture

```{r expand_expr_7, echo=TRUE, eval=TRUE, error=TRUE}
mtcars_tbl %>% group_by_mean(group_var = cyl, mean_var = hp)
mtcars_tbl %>% group_by_mean(group_var = gear, mean_var = mpg)
```

## Expression Capture Summary

> - We can pass expressions to functions.

> - We can capture the expressions in order to manipulate them.

> - We can use functions/operators from the `rlang` package such as `enexpr`, `!!`, `as_string`, and `:=` to quote (capture), manipulate, and unquote expressions passed to a function.


# UDS 3 Data

## UDS 3 Data

Why would metaprogramming be useful for managing UDS 3 data?

Sample dataset using [REDCap Collaborative UDS 3.0](https://www.alz.washington.edu/NONMEMBER/FALL17/Data/Hunt.pdf) data dictionary from **KU ADC**.

```{r uds3_data_1, results="asis"}
options(knitr.kable.NA = '')
dataset_1 <- 
  tibble(ptid = c("PT01", 
                  "PT02", "PT02",
                  "PT03", "PT03", "PT03"),
         packet = c("I",
                    "I", "F",
                    "I", "F", "T"),
         visitmo = c(rep(1L, 1),
                     rep(2L, 2),
                     rep(3L, 3)),
         visitday = c(rep(1L, 1),
                      rep(2L, 2),
                      rep(3L, 3)),
         visityr = c("2015",
                     "2015", "2016",
                     "2015", "2016", "2017"),
         # race = c(3L, 
         #          50L, NA_integer_,
         #          4L, NA_integer_, NA_integer_),
         sex = c(2L, 
                 2L, NA_integer_,  
                 1L, NA_integer_, NA_integer_),
         fu_sex = c(NA_integer_, 
                    NA_integer_, 2L,
                    NA_integer_, 1L, NA_integer_),
         tele_sex = c(NA_integer_, 
                      NA_integer_, NA_integer_,
                      NA_integer_, NA_integer_, 1L),
  ) 
dataset_1 %>% 
  pretty_print()
```

## UDS 3 Data (2)

Why would metaprogramming be useful for manage UDS 3 data?

_Another_ sample dataset from Form A3. Here participant siblings' data can make up 20 possible siblings (`sib1`-`sib20`) by 7 fields (`mob`, `yob`, `agd`, `neu`, `pdx`, `moe`, `ago`) on three 3 forms (I, F, T): 20 x 7 x 3 = 420 fields.

```{r uds3_data_2, results="asis"}
options(knitr.kable.NA = '')
dataset_2 <-
  tibble(ptid = c("PT01", 
                  "PT02", "PT02",
                  "PT03", "PT03", "PT03"),
         packet = c("I",
                    "I", "F",
                    "I", "F", "T"),
         sib1mob = c(1L,
                     2L, NA_integer_,
                     3L, NA_integer_, NA_integer_),
         sib1yob = c(1941L,
                     1942L, NA_integer_,
                     1943L, NA_integer_, NA_integer_),
         sib1agd = c(NA_integer_,
                     NA_integer_, NA_integer_,
                     72L, NA_integer_, NA_integer_),
         # `_` = rep("...", 6),
         fu_sib1mob = c(NA_integer_,
                        NA_integer_, 2L,
                        NA_integer_, 3L, NA_integer_),
         fu_sib1yob = c(NA_integer_,
                        NA_integer_, 1942L, 
                        NA_integer_, 1943L, NA_integer_),
         fu_sib1agd = c(NA_integer_,
                        NA_integer_, NA_integer_,
                        NA_integer_, 72L, NA_integer_),
         # `__` = rep("...", 6),
         tele_sib1mob = c(NA_integer_,
                          NA_integer_, NA_integer_,
                          NA_integer_, NA_integer_, 3L),
         tele_sib1yob = c(NA_integer_,
                          NA_integer_, NA_integer_,
                          NA_integer_, NA_integer_, 1943L),
         tele_sib1agd = c(NA_integer_,
                          NA_integer_, NA_integer_,
                          NA_integer_, NA_integer_, 72L),
         # `___` = rep("...", 6),
  ) 
dataset_2 %>% 
  pretty_print_scroll()
```

## Coalescing Sparsity - Dataset 1

```{r coal_sparse_d1_1}
dataset_1 %>% 
  pretty_print()
```

## Coalescing Sparsity - Dataset 1

Use `dplyr::coalesce`.

```{r coal_sparse_d1_2a, echo=TRUE, eval=FALSE}
dataset_1 %>% 
  mutate(sex = coalesce(sex, fu_sex, tele_sex))
```

```{r coal_sparse_d1_2b, echo=FALSE, eval=TRUE}
dataset_1 %>% 
  mutate(sex = coalesce(sex, fu_sex, tele_sex)) %>% 
  pretty_print()
```

## Coalescing Sparsity - Dataset 1

Remove redundant fields with `dplyr::select`.

```{r coal_sparse_d1_3a, echo=TRUE, eval=FALSE}
dataset_1 %>% 
  mutate(sex = coalesce(sex, fu_sex, tele_sex)) %>% 
  select(-fu_sex, -tele_sex)
```

```{r coal_sparse_d1_3b, echo=FALSE, eval=TRUE}
dataset_1 %>% 
  mutate(sex = coalesce(sex, fu_sex, tele_sex)) %>% 
  select(-fu_sex, -tele_sex) %>% 
  pretty_print()
```

## Coalescing Sparsity - Dataset 2

Easy enough with one field represented in initial, follow-up, and telephone forms.

What about our second dataset from Form A3?

```{r coal_sparse_d2_1, echo=FALSE, eval=TRUE}
dataset_2 %>% 
  pretty_print_scroll()
```

## Coalescing Sparsity - Dataset 2

Brute force approach.

```{r coal_sparse_d2_2a, echo=TRUE, eval=FALSE}
dataset_2 %>% 
  mutate(sib1mob = coalesce(sib1mob, fu_sib1mob, tele_sib1mob),
         sib1yob = coalesce(sib1yob, fu_sib1yob, tele_sib1yob),
         sib1agd = coalesce(sib1agd, fu_sib1agd, tele_sib1agd)) %>% 
  select(-fu_sib1mob, -tele_sib1mob, -fu_sib1yob, -tele_sib1yob,
         -fu_sib1agd, -tele_sib1agd)
```

```{r coal_sparse_d2_2b, echo=FALSE, eval=TRUE}
(dataset_2_out <- dataset_2 %>% 
   mutate(sib1mob = coalesce(sib1mob, fu_sib1mob, tele_sib1mob),
          sib1yob = coalesce(sib1yob, fu_sib1yob, tele_sib1yob),
          sib1agd = coalesce(sib1agd, fu_sib1agd, tele_sib1agd)) %>% 
   select(-fu_sib1mob, -tele_sib1mob, -fu_sib1yob, -tele_sib1yob,
          -fu_sib1agd, -tele_sib1agd)) %>% 
  pretty_print_scroll()
```

## Coalescing Sparsity - Dataset 2

Brute force will be daunting once we start including the 19 other possible siblings (`sib2`--`sib20`) and the 4 other fields (`neu`, `pdx`, `moe`, `ago`).

Tidy Eval to the rescue!

## Coalescing Sparsity - Dataset 2

First we need a function that finds only the initial visit field names.

For example, in `dataset_2`...

> - `ptid`, `packet` are visit-packet independent.

> - `sib1mob`, `sib1yob`, `sib1agd` is from the initial visit packet.

> - `fu_sib1mob`, `fu_sib1yob`, `fu_sib1agd` is from the follow-up visit packet.

> - `tele_sib1mob`, `tele_sib1yob`, `tele_sib1agd` is from the telephone visit packet.

## Coalescing Sparsity - Dataset 2

So, given the field names

`ptid`, `packet`, <br/>
`sib1mob`, `sib1yob`, `sib1agd`, <br/>
`fu_sib1mob`, `fu_sib1yob`, `fu_sib1agd`, </br> 
`tele_sib1mob`, `tele_sib1yob`, `tele_sib1agd`

... we'd like a function that returns

`sib1mob`, `sib1yob`, `sib1agd`

## Coalescing Sparsity - Dataset 2

```{r coal_sparse_d2_3, echo=TRUE, eval=TRUE}
reduce_ift_fieldnames <- function(ift_fields) {
  
  # Locate fields beginning with `fu_` or `tele_`
  ft_fields_loc <- str_detect(ift_fields, "^fu_|^tele_")
  
  # Return only `fu_` or `tele_` fields
  ft_fields <- ift_fields[ft_fields_loc]
  
  # Strip off `fu_` or `tele_` to get initial visit field names
  i_fields <- str_replace_all(ft_fields, 
                              pattern = "^fu_|^tele_", 
                              replacement = "")
  
  # There'll be repeats, so return only unique field names
  unique(i_fields)
}
```

## Coalescing Sparsity - Dataset 2

Input:

```{r coal_sparse_d2_4a, echo=TRUE, eval=TRUE}
names(dataset_2)
```

Output:

```{r coal_sparse_d2_4b, echo=TRUE, eval=TRUE}
reduce_ift_fieldnames(names(dataset_2))
```

## Coalescing Sparsity - Dataset 2

With this helper function `reduce_ift_fieldnames` in place, we can create a function that uses Tidy Eval to coalesce all initial visit, follow-up visit, and telephone visit fields.

We'll call this new function `coalesce_all_ift_fields`.

But let's first review the input and desired output.

## Coalescing Sparsity - Dataset 2

Input: `dataset_2`

```{r coal_sparse_d2_5a, echo=FALSE, eval=TRUE}
dataset_2 %>% pretty_print_scroll()
```

## Coalescing Sparsity - Dataset 2

Output:

```{r coal_sparse_d2_5b, echo=FALSE, eval=TRUE}
dataset_2_out %>% pretty_print_scroll()
```

## Coalescing Sparsity - Dataset 2

```{r coal_sparse_d2_6, echo=TRUE, eval=TRUE}
coalesce_all_ift_fields <- function(df) {
  # Get initial visit field names, follow-up names, and telephone names
  i_fields <- reduce_ift_fieldnames(names(df))
  f_fields <- paste0("fu_", i_fields)
  t_fields <- paste0("tele_", i_fields)
  
  # Iterate over each of the `i_fields`, applying `coalesce` and `select`
  for (n in seq_along(i_fields)) { # n = 1, 2, 3, 4, ...
    df <- df %>%
      mutate( !!i_fields[n] := coalesce(df[[ i_fields[n] ]],
                                        df[[ f_fields[n] ]],
                                        df[[ t_fields[n] ]] )) %>%
      select( -!!f_fields[[n]], -!!t_fields[[n]] )
  }
  # Return the dataframe
  df
}
```

## Coalescing Sparsity - Dataset 2

```{r coal_sparse_d2_7a, echo=TRUE, eval=FALSE}
dataset_2 %>% coalesce_all_ift_fields()
```

```{r coal_sparse_d2_7b, echo=FALSE, eval=TRUE}
dataset_2 %>% coalesce_all_ift_fields() %>% pretty_print_scroll()
```

## Coalescing Sparsity - Dataset 1

Let's test this new function on the first simple dataset.

Input: `dataset_1`

```{r}
dataset_1 %>% pretty_print()
```

## Coalescing Sparsity - Dataset 1

And...

Output: `coalesce_all_ift_fields(dataset_1)`

```{r}
coalesce_all_ift_fields(dataset_1) %>% pretty_print()
```














